---
title: 'Exploratory analysis: Urinary biomarkers for pancreatic cancer'
author: "Cristina Juárez Alía"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r include=FALSE}
library(dplyr)
library(tidyverse)
library(RColorBrewer)
library(reshape2)
install.packages("car")
renv::snapshot()
```

```{r include=FALSE}
# Importing data from .csv (original dataset)
data <- read.csv("C:/Users/cjuar/Desktop/Projects/Urinary-biomarkers-for-pancreatic-cancer/data/Debernardi et al 2020 data.csv", sep= ",", header = TRUE)
```

# Overview of initial dataset

The dataset belongs to the study of Debernardi S. (2018) and can be downloaded from Kaggle (https://www.kaggle.com/datasets/johnjdavisiv/urinary-biomarkers-for-pancreatic-cancer).

## Dimensions

```{r dim}
dim(data)
```

## Structure

```{r dim}
str(data)
```

The variables `patient_cohort`, `sample_origin`, `sex`, `diagnosis`, `stage`, `benign_sample_diagnosis` should be coerced to factors later (preprocessing).

## Summary

```{r summary}
summary(data)
```
The NA's in `REG1A` and `plasma_CA19_9` correspond to unmeasured values. Only 350 and 306 patients, respectively, were measured these parameters. This information is provided in the documentation provided by the authors. Therefore, since these NA's are established by design (authors's decision), we won't conduct a missing data analysis.

# Preprocessing

## Handling of insconsistencies 

`benign_sample_diagnosis` is a character vector so it is likely that there are some typos. 

```{r echo=FALSE}
unique(data$benign_sample_diagnosis)
```

```{r typos}
# Elimination of white spaces
data$benign_sample_diagnosis <- trimws(data$benign_sample_diagnosis)

# Correction of some typos
# NOTE: These typos have been detected manually.
data$benign_sample_diagnosis <- gsub("Pseuodcyst", "Pseudocyst", data$benign_sample_diagnosis)
data$benign_sample_diagnosis <- gsub("Cholesterolsis", "Cholesterosis", data$benign_sample_diagnosis)
data$benign_sample_diagnosis <- gsub("Choledocholiathiasis", "Choledocholithiasis", data$benign_sample_diagnosis)
```


## Coercion 

```{r coercion}
# Before coercing variable to factors we change some of the values for better rediability of the code

data <- data %>% mutate(
  
  diagnosis = case_when(diagnosis == 1 ~ "control",
                        diagnosis == 2 ~ "benign",
                        diagnosis == 3 ~ "PDAC"),
  
  stage = case_when(stage == "" ~ NA_character_,
                    TRUE ~ stage),
  
  benign_sample_diagnosis = case_when(benign_sample_diagnosis == "" ~ NA_character_,
                                      TRUE ~ benign_sample_diagnosis),
  
  across(c(patient_cohort, sample_origin, sex, diagnosis, stage, benign_sample_diagnosis), as.factor))
```

---

Finally, we save the preprocessed dataset, ready to work with:

```{r saving}
save(data, file="data/df.rda")
```

# Exploratory analysis

## Contingency table: Sex vs Diagnostic Group

```{r}
sex_vs_diagnosis <- table(data$sex, data$diagnosis)
```

### Chi-squared test: Association between Sex and Diagnostic Group

We can complement with a Chi-squared test, which evaluates if there is an association between sex vs diagnosis.

```{r}
chisq.test(sex_vs_diagnosis)
```

According to the test, there is an association between sex and diagnosis. 
This should be accounted while interpreting the results.


## Multivariate analysis

### Violin Plots: Age by Diagnosis

```{r}
ggplot(data = data, aes(x = diagnosis, y = age, fill = diagnosis)) +
  geom_violin(trim = FALSE) +
  #coord_flip() # Flip the plot the other way around
  #stat_summary(fun = mean, geom = "point", shape = 21, colour = "blue") + # mean
  geom_boxplot(width = 0.1) +  # Adding boxplots
  scale_fill_brewer(palette="Accent") +
  labs(title = "Age by Diagnosis") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) 
```


#### Checking Normality and Variance of Age across Diagnostic groups

##### Normality test: Shapiro-Wilk

```{r}
# Doing the test for each diagnosis group:
by(data$age, data$diagnosis, shapiro.test)
```
There is sufficient evidence to reject the assumption of normality in the `benign` and `PDAC` groups based on the Shapiro-Wilk test (p < 0.05). *Therefore, we won't use ANOVA*

##### Variance test: Levene's Test

NOTE: We don't use Barlett's Test because groups `benign` and `PDAC` deviate from normality.

```{r}

leveneTest(age ~ diagnosis, data = data)
```

*Unfinished!*

# Bibliography 

Debernardi S, O’Brien H, Algahmdi AS, Malats N, Stewart GD, et al. (2020) A combination of urinary biomarker panel and PancRISK score for earlier detection of pancreatic cancer: A case–control study. PLOS Medicine 17(12): e1003489. https://doi.org/10.1371/journal.pmed.1003489
